Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-21 paper/main.tex
checking in updates to methods

In the set of analyses reported in Figure~\ref{fig:predictions}, we assessed whether estimates of participants' knowledge at the embedding coordinates of individual quiz questions could reliably predict their ability to correctly answer those questions.
In essence, we treated each question a given participant answered on a given quiz as a ``lecture'' consisting of a single timepoint, and used Equation~\ref{eqn:prop} to estimate the participant's knowledge for its embedding coordinate based on their performance on all \textit{other} questions they answered on that same quiz (``All questions''; Fig.~\ref{fig:predictions}, top row). 
Additionally, for each lecture-related question (i.e., excluding questions about general physics knowledge), we computed analogous knowledge estimates based on all other questions the participant answered on the same quiz about (1) the same lecture as the target question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows), and (2) the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).
%For each lecture-related question (i.e., excluding questions about general physics knowledge), we also computed analogous knowledge estimates based on two subsets of the other questions the participant answered on the same quiz: (1) all other questions about the same lecture as the target question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows), and (2) all questions about the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).
%For each lecture-related question (i.e., excluding questions about general physics knowledge), we also computed analogous knowledge estimates based on:~(1) all other questions the participant answered on the same quiz about the same lecture as the target question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows), and (2) all questions they answered on the same quiz about the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).
%We also computed analogous knowledge estimates for each lecture-related question based on (1) all other questions answered by the same participant, on the same quiz about the same lecture as the target question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows) and (2) all questions from the same participant and quiz about the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).

In each version of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and separately for each of the three quizzes (i.e., column in Fig.~\ref{fig:predictions}),  we then fit a generalized linear mixed model (GLMM) with a logistic link function to the set of knowledge estimates for all questions that participants answered on the given quiz.
We implemented these models in \texttt{R} using the \texttt{lme4} package~\citep{BateEtal15a} and fit them following guidance from~\citet{BateEtal15b} and \citet{MatuEtal17}. Specifically, we initially fit each model with the maximal random effects structure afforded our data, which we identified as:
\[
    \mathtt{accuracy \sim knowledge + (knowledge\ \vert\ participant) + (knowledge\ \vert\ question)}
\]
Here, ``\texttt{accuracy}'' is a binary value indicating whether each target question was answered correctly or incorrectly, ``\texttt{knowledge}'' is estimated knowledge at each target question's embedding coordinate, ``\texttt{participant}'' is a unique identifier assigned to each participant, and ``\texttt{question}'' is a unique identifier assigned to each quiz question.
For models we fit using knowledge estimates for target questions about multiple content areas (i.e., in the ``All questions'' version of the analysis), we also included an additional random effect term, $\mathtt{(knowledge\ \vert\ lecture)}$, where ``\texttt{lecture}'' is a categorical value denoting whether the target question was about \textit{Four Fundamental Forces}, \textit{Birth of Stars}, or general physics knowledge.
Note that with our coding scheme, identifiers for each \texttt{question} are implicitly nested within levels of \texttt{lecture} and do not require explicit nesting in our model formula.

When necessary, we then iteratively removed random effects from the maximal model until it successfully converged with a full rank (i.e., non-singular) random effects variance-covariance matrix. 

@@

When this required eliminating multiple terms whose estimates reached the boundary of their parameter space (i.e., variance components of 0 or correlation terms of  $\pm 1$), we found (qualitatively) that the order in which we did so typically did not change the set of terms that needed to be removed in order for the model to converge to a non-degenerate solution.
However, in order to ensure we fit all of the models for these analyses in a consistent way, we devised some simple heuristics for choosing which boundary-estimated parameter to eliminate in each stepwise reduction of the model's complexity, based on a combination of popular approaches~\citep{BarrEtal13,BateEtal15b,MatuEtal17}.

@@

Finally, once the model successfully converged with a full rank covariance matrix, we did not remove random effects for any other reasons (e.g., to lowered AIC, increase log-likelihood, etc.). 

@@

To assess the predictive value of our knowledge estimates for individual quiz questions, we used the same sets of observations used to fit each GLMM to fit a second set of ``null'' models (similarly with a logistic link function). We fit these models with the formula:
\[
    \mathtt{accuracy \sim (1\ \vert\ participant) + (1\ \vert\ question)}
\]

>>>>

In the set of analyses reported in Figure~\ref{fig:predictions}, we assessed whether estimates of participants' knowledge at the embedding coordinates of individual quiz questions could be used to reliably predict their ability to correctly answer those questions.
In essence, we treated each question a given participant answered on a given quiz as a ``lecture'' consisting of a single timepoint, and used Equation~\ref{eqn:prop} to estimate the participant's knowledge for its embedding coordinate based on their performance on all \textit{other} questions they answered on that same quiz (``All questions''; Fig.~\ref{fig:predictions}, top row). 
Additionally, for each lecture-related question (i.e., excluding questions about general physics knowledge), we computed analogous knowledge estimates based on all other questions the participant answered on the same quiz about (1) the same lecture as the target question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows), and (2) the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).
%For each lecture-related question (i.e., excluding questions about general physics knowledge), we also computed analogous knowledge estimates based on two subsets of the other questions the participant answered on the same quiz: (1) all other questions about the same lecture as the target question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows), and (2) all questions about the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).
%For each lecture-related question (i.e., excluding questions about general physics knowledge), we also computed analogous knowledge estimates based on:~(1) all other questions the participant answered on the same quiz about the same lecture as the target question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows), and (2) all questions they answered on the same quiz about the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).
%We also computed analogous knowledge estimates for each lecture-related question based on (1) all other questions answered by the same participant, on the same quiz about the same lecture as the target question (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows) and (2) all questions from the same participant and quiz about the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).

In each version of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and separately for each of the three quizzes (i.e., column in Fig.~\ref{fig:predictions}),  we then fit a generalized linear mixed model (GLMM) with a logistic link function to the set of knowledge estimates for all questions that participants answered on the given quiz.
We implemented these models in \texttt{R} using the \texttt{lme4} package~\citep{BateEtal15a} and fit them following guidance from~\citet{BateEtal15b} and \citet{MatuEtal17}. Specifically, we initially fit each model with the maximal random effects structure afforded by our design, which we identified as:
\[
    \mathtt{accuracy \sim knowledge + (knowledge\ \vert\ participant) + (knowledge\ \vert\ question)}
\]
Here, ``\texttt{accuracy}'' is a binary value indicating whether each target question was answered correctly or incorrectly, ``\texttt{knowledge}'' is estimated knowledge at each target question's embedding coordinate, ``\texttt{participant}'' is a unique identifier assigned to each participant, and ``\texttt{question}'' is a unique identifier assigned to each quiz question.
For models we fit using knowledge estimates for target questions about multiple content areas (i.e., in the ``All questions'' version of the analysis), we also included an additional random effect term, $\mathtt{(knowledge\ \vert\ lecture)}$, where ``\texttt{lecture}'' is a categorical value denoting whether the target question was about \textit{Four Fundamental Forces}, \textit{Birth of Stars}, or general physics knowledge.
Note that with our coding scheme, identifiers for each \texttt{question} are implicitly nested within levels of \texttt{lecture} and do not require explicit nesting in our model formula.

We then iteratively removed random effects from the maximal model until it successfully converged with a full rank (i.e., non-singular) random effects variance-covariance matrix. 

@@

When this required eliminating multiple terms whose estimates reached the boundary of their parameter space (i.e., variance components of 0 or correlation terms of  $\pm 1$), we found that the order in which we did so typically did not change the set of terms that needed to be removed in order for the model to converge to a non-degenerate solution.
However, in order to ensure we fit all models used in these analyses in a consistent way, we devised some simple heuristics for choosing which boundary-estimated parameter to eliminate in each stepwise reduction of the model's complexity, based on a combination of popular approaches~\citep{BarrEtal13,BateEtal15b,MatuEtal17}.

@@

Finally, once the model successfully converged with a full rank covariance matrix, we did not remove random effects for any other reasons (e.g., to lower AIC, increase log-likelihood, etc.). 

@@

%To assess the predictive value of our knowledge estimates for individual quiz questions, we used the same sets of observations used to fit each GLMM to fit a second set of ``null'' models (similarly with a logistic link function). We fit these models with the formula:
To assess the predictive value of our knowledge estimates, we compared each GLMM's ability to discriminate between correctly and incorrectly answered questions to that of an analogous model that did not consider estimated knowledge. Specifically, we used the same sets of observations with which we fit each ``full'' model to fit a second ``null'' model, with the formula:
\[
    \mathtt{accuracy \sim (1\ \vert\ participant) + (1\ \vert\ question)}
\]
